Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-01-22 Manuscript/HI_Model_ms.Rmd
updated manuscript files

The State of Hawaiʻi exemplifies the challenges of projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances, with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016].

We used a fully coupled consider four unique future scenarios

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawai"i LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

>>>>

The State of Hawaiʻi exemplifies the challenges associated with projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances, with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016]. Although these past trends surely inform the future impact of climate and land-use change on ecosystem carbon balance, high spatial and temporal heterogeneity complicates realistic projection efforts. To date only one study has attempted to integrate land-use, climate, and natural disturbances into future projections of Hawaiian ecosystem carbon balance, with projections limited to the mid-21st century under a single land-use change scenario and moderate radiative forcing [SRES A1B, equivalent to RCP 6; @selmants_baseline_2017].

In this study, we consider four unique scenarios exploring a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions. We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands under each of these four scenarios on an annual basis for the period 2010–2100 [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. The four unique scenarios represent all combinations of two land-use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and RCP 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary over time in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. The goals of this study were to (a) estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, (b) quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and (c) assess the sensitivity of model estimates to the introduction of a CFE on NPP.

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### 2021-01-22 Manuscript/HI_Model_supp.Rmd
updated manuscript files

---
title: "Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change"
bibliography: HI_Model.bib
csl: iop.csl
fontsize: 12pt
geometry: margin=1in
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
output: 
  pdf_document: 
    latex_engine: xelatex
urlcolor: blue

@@

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
Paul C. Selmants^1,6^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,  
Parker C. Trauernicht^3^, Abby G. Frazier^4^, Gregory P. Asner^5^

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA   
^2^U.S. Geological Survey, Seattle, WA, USA  
^3^University of Hawai`i at Mānoa, Honolulu, HI, USA  
^4^The East-West Center, Honolulu, HI, USA  
^5^Arizona State University, Tempe, AZ, USA  
^6^Author to whom correspondence should be addressed

>>>>

---
title: "Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change"
bibliography: HI_Model.bib
csl: iop.csl
geometry: margin=1in
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
output: 
  pdf_document: 
    latex_engine: xelatex
mainfont: Times New Roman
fontsize: 12pt
urlcolor: blue

@@

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
\centering
Paul C. Selmants^1,6^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,  
Parker C. Trauernicht^3^, Abby G. Frazier^4^, Gregory P. Asner^5^

\raggedright
**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA   
^2^U.S. Geological Survey, Seattle, WA, USA  
^3^University of Hawaiʻi at Mānoa, Honolulu, HI, USA  
^4^The East-West Center, Honolulu, HI, USA  
^5^Arizona State University, Tempe, AZ, USA  
^6^Author to whom correspondence should be addressed
